---
title: "politicalanalyticsfinal_projectfirstvis"
author: "vat2129"
date: "2025-11-19"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries & Set Up
```{r}
library(shiny)
library(dplyr)
library(foreign)
library(tidyr)
library(ggplot2)
library(maps)
library(stringr)
library(usmap)
library(gganimate)
library(gifski)
```
Cleaning & Exploring the Data
```{r}
bigclimatedata <- foreign::read.spss("~/datasci01/data/CCAM SPSS Data 2008-2024.sav", to.data.frame=TRUE)

#wanted to understand what % of each variable was NA
total_rows <- nrow(bigclimatedata)
na_counts <- colSums(is.na(bigclimatedata))
percentage_na <- (na_counts / total_rows) * 100
percentage_na

#here we are funneling the data into the pieces that will be pertinent to our analysis
climatedata <- bigclimatedata %>% 
  select(case_ID, wave, weight_wave, weight_aggregate, year, happening, worry, 
         reg_CO2_pollutant, fund_research, party_w_leaners, region9, educ, 
         educ_category, age, age_category, income, income_category, generation,
         starts_with("harm"), cause_recoded) %>% 
  mutate(party_w_leaners = case_when(
    party_w_leaners == "Refused" ~ NA,
    party_w_leaners == "No party/Not interested in politics" ~ "Independent/Other",
    TRUE ~ party_w_leaners)) %>%
  mutate(across(starts_with("harm"), ~case_when(
    . == "Don't know" ~ NA, 
    . == "Not at all" ~ 1,
    . == "Only a little" ~ 2,
    . == "A moderate amount" ~ 3,
    . == "A great deal" ~ 4,
    . == "Refused" ~ NA))) %>%
  mutate(worry = case_when(
    worry == "Not at all worried" ~ 1,
    worry == "Not very worried" ~ 2,
    worry == "Somewhat worried" ~ 3,
    worry == "Very worried" ~ 4,
    worry == "Refused" ~ NA)) %>%
  mutate(harm_worry = worry) %>%
  mutate(non_na_count = rowSums(!is.na(across(starts_with("harm"))))) %>% 
  mutate(concern_score = rowMeans(across(starts_with("harm")), na.rm = TRUE))

```
Numbers to Consider
```{r}
#2256 people in this data set explicitly believe that global warming is not happening
climatedata %>% 
  select(cause_recoded) %>%
  count(cause_recoded)

concern_trends <- climatedata %>%
  group_by(year, region9) %>%
  summarise(
    #calulate a weighted average
    mean_concern_score = weighted.mean(concern_score, w = weight_aggregate, na.rm = TRUE),
    n = n(),
    .groups = 'drop') %>%
  #remove NA 
  filter(!is.na(region9))
concern_trends

#wanted to understand the concern across different generations
climatedata %>%
  summarize(mean = mean(concern_score, na.rm = TRUE), 
            sd = sd(concern_score, na.rm = TRUE),
            n = n(),
            .by = c(party_w_leaners, generation))%>%
  arrange(party_w_leaners, generation)

#wanted to understand the concern across different regions
climatedata %>%
  summarize(mean = mean(concern_score, na.rm = TRUE), 
            sd = sd(concern_score, na.rm = TRUE),
            n = n(),
            .by = c(party_w_leaners, region9))%>%
  arrange(party_w_leaners, region9)
```

FIGURE 1 - Climate Concern by Region and Party 
```{r}
#FIGURE 1 - Climate Concern by Region and Party 

#here is a more general look of the various parties concern of climate change (not factoring in time)
region_party_scores <- climatedata %>%
  group_by(region9, party_w_leaners) %>%
  summarize(mean_concern = weighted.mean(concern_score, w = weight_aggregate, na.rm = TRUE), 
    .groups = 'drop') %>%
  filter(!is.na(party_w_leaners))
  

region_states <- tibble(
  state = state.abb,
  region9 = case_when(
    state %in% c("CT","MA","ME","NH","RI","VT") ~ "New England",
    state %in% c("NJ","NY","PA") ~ "Mid-Atlantic",
    state %in% c("IL","IN","MI","OH","WI") ~ "East-North Central",
    state %in% c("IA","KS","MN","MO","ND","NE","SD") ~ "West-North Central",
    state %in% c("DC","DE","FL","GA","MD","NC","SC","VA","WV") ~ "South Atlantic",
    state %in% c("AL","KY","MS","TN") ~ "East-South Central",
    state %in% c("AR","LA","OK","TX") ~ "West-South Central",
    state %in% c("AZ","CO","ID","MT","NM","NV","UT","WY") ~ "Mountain",
    state %in% c("AK","CA","HI","OR","WA") ~ "Pacific",
    TRUE ~ NA))

map_df <- region_states %>%
  left_join(region_party_scores, by = "region9")


plot_usmap(data = map_df,
  values = "mean_concern",
  color = "white") +
  scale_fill_continuous(name = "Climate Concern",
                        label = scales::comma) +
  labs(title = "Climate Concern by Region and Party",
       subtitle = "Mean concern score, faceted by party") +
  facet_wrap(~ party_w_leaners) +
  theme(legend.position = "right")
```

Here we are visualizing climate concern which is an index that we created utilizing six survey questions from the data set. The highest concern score someone could achieve is 4, meaning that were very worried about climate change, while the lowest score possible would be a 1 which would mean you were not concerned about climate change in any way. In Figure 1 we are visualize concern by region of the country separated out by party. For this visual we took the previous cleaned data (climate data) that included our questions of interest and demographic information, and that created the new climate concern index. 
We then took this cleaned data and grouped it by region and party, as well as created a weighted mean with the aggregate weights including in the data set. We also filtered out any NA values. When we wanted to take a closer look at region, the only visual that felt appropriate was a map. As visible in the table above, even dividing by region and party we do not have a subsection with n less than 366 (Independents in New England). What we visualize here aligns very closely with what our hypothesis was. Party seemed to be a much stronger indicator of climate concern than region with Democrats being most concerned, then Independents, and then Republicans. There was only very slight concern variations by region.

FIGURE 2 - Evolution of Climate Concern by U.S. Region
```{r}
#FIGURE 2 - Evolution of Climate Concern by U.S. Region

#here is the general map over time (all parties)
region_scores_time <- climatedata %>%
  mutate(year = as.integer(year)) %>%
  group_by(year, region9) %>%
  #calculate the weighted mean concern score
  summarize(mean_concern = weighted.mean(concern_score, w = weight_aggregate, na.rm = TRUE), 
    .groups = 'drop') %>%
  complete(year, region9) %>%
  #remove rows where a region/year combination resulted in NA mean concern
  filter(!is.na(mean_concern))

region_states <- tibble(state = state.abb,
  region9 = case_when(state %in% c("CT","MA","ME","NH","RI","VT") ~ "New England",
    state %in% c("NJ","NY","PA") ~ "Mid-Atlantic",
    state %in% c("IL","IN","MI","OH","WI") ~ "East-North Central",
    state %in% c("IA","KS","MN","MO","ND","NE","SD") ~ "West-North Central",
    state %in% c("DC","DE","FL","GA","MD","NC","SC","VA","WV") ~ "South Atlantic",
    state %in% c("AL","KY","MS","TN") ~ "East-South Central",
    state %in% c("AR","LA","OK","TX") ~ "West-South Central",
    state %in% c("AZ","CO","ID","MT","NM","NV","UT","WY") ~ "Mountain",
    state %in% c("AK","CA","HI","OR","WA") ~ "Pacific",
    TRUE ~ NA_character_))

map_df_time <- region_states %>%
  left_join(region_scores_time, by = "region9", relationship = "many-to-many") %>%
  filter(!is.na(region9))

#the base plot for the US map
base_map_plot <- plot_usmap(data = map_df_time, values = "mean_concern", color = "white") +
  
  #construct the fill scale
  scale_fill_continuous(
    name = "Mean Climate Concern Score",
    low = "lightblue",
    high = "darkred",
    limits = c(min(map_df_time$mean_concern, na.rm = TRUE), 
               max(map_df_time$mean_concern, na.rm = TRUE))) +
  #labels and title
  labs(title = "Evolution of Climate Concern by U.S. Region",
    #so the graph shows the wave its showing 
    subtitle = "Year: {closest_state}") +
  #make sure it is clearly legible
  theme(legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, face = "bold"))

#put it together
animated_map_plot <- base_map_plot +
  #this is what will create the cycling of the year
  transition_states(year, transition_length = 2, state_length = 1)

#animate (commented out for pdf)
#animate(animated_map_plot, renderer = gifski_renderer(), nframes = 50, fps = 5, 
       # width = 800, height = 500)
```

FIGURE 3,4,5: Evolution of Climate Concern by U.S. Region by Party 
```{r}
#as we were interested in understanding how concern changed by region across time, we created a visual that will show a map of the US each year that the survey ran and display the concern by party
make_party_animation <- function(party_name) {

  #to create the visual we had to create a table such that the map fcn could understand the regions
  region_states <- tibble(state = state.abb,
   region9 = case_when(state %in% c("CT","MA","ME","NH","RI","VT") ~ "New England",
    state %in% c("NJ","NY","PA") ~ "Mid-Atlantic",
    state %in% c("IL","IN","MI","OH","WI") ~ "East-North Central",
    state %in% c("IA","KS","MN","MO","ND","NE","SD") ~ "West-North Central",
    state %in% c("DC","DE","FL","GA","MD","NC","SC","VA","WV") ~ "South Atlantic",
    state %in% c("AL","KY","MS","TN") ~ "East-South Central",
    state %in% c("AR","LA","OK","TX") ~ "West-South Central",
    state %in% c("AZ","CO","ID","MT","NM","NV","UT","WY") ~ "Mountain",
    state %in% c("AK","CA","HI","OR","WA") ~ "Pacific",
    TRUE ~ NA_character_))
  
  #we filter to be able to seperate by party
  party_data <- climatedata %>%
    filter(party_w_leaners == party_name) %>%
    mutate(year = as.integer(year)) %>%
    group_by(year, region9) %>%
    summarize(mean_concern = weighted.mean(concern_score, w = weight_aggregate, na.rm = TRUE),
              .groups = "drop")

  #join our states data with our party data
  party_map <- region_states %>%
    left_join(party_data, by = "region9", relationship = "many-to-many") %>%
    filter(!is.na(mean_concern))

  #fix the color scale 
  fill_limits <- range(party_mapf$mean_concern, na.rm = TRUE)

  #base map
  base_map <- plot_usmap( data = party_map, values = "mean_concern", color = "white") +
    scale_fill_continuous(name = "Mean Climate Concern Score",
      low = "lightblue", high = "darkred", limits = fill_limits) 
  + labs(title = paste0("Climate Concern by U.S. Region — ", party_name),
      subtitle = "Year: {frame_time}") 
  + theme(legend.position = "right", plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, face = "bold"))

  #animation details
  animated_plot <- base_map + transition_time(year) + ease_aes("linear")

  animate(animated_plot, renderer = gifski_renderer(), nframes = 80, fps = 6,
          width = 800, height = 500, end_pause = 10)
}

#Figures 3, 4, & 5 Evolution of Climate Concern by U.S. Region by Party 
#anim_dem <- make_party_animation("Democrats")
#anim_rep <- make_party_animation("Republicans")
#anim_ind <- make_party_animation("Independent/Other")

```

